# Predict parental breeding values


```{r setup, include=F, echo=F}
library(tidyverse); library(genomicMateSelectR); library(sommer); library(gt)
```

Now that we tested genomic prediction accuracy using cross-validation, we can run genomic predictions.

In the previous section where we [introduced genomic prediction](intro-to-genomic-prediction), we learned how to use the `mmer()` function in `library(sommer)` to run GBLUP models and also rrBLUP models.

For the actual predictions, we can use the function build into `library(genomicMateSelectR)`, `runGenomicPredictions()`. You can find the documentation for that function by [clicking here](https://wolfemd.github.io/genomicMateSelectR/reference/runGenomicPredictions.html).

`runGenomicPredictions()` is a wrapper that uses `mmer()` under-the-hood. It expects de-regressed BLUPs and weights as input.


## Set-up for the predictions

Similar set-up to what we did for cross-validation.

Load the BLUps and the kinship matrix.
```{r}
blups<-readRDS(here::here("output","blups.rds"))
A<-readRDS(file=here::here("output","kinship_add.rds"))
```

```{r}
blups %<>% 
     # based on cross-validation, decided to exclude MCMDS from this analysis
     filter(Trait != "MCMDS") %>% 
     # need to rename the "blups" list to comply with the runCrossVal function
     rename(TrainingData=blups) %>% 
     dplyr::select(Trait,TrainingData) %>% 
     # need also to remove phenotyped-but-not-genotyped lines
     mutate(TrainingData=map(TrainingData,
                             ~filter(.,germplasmName %in% rownames(A)) %>% 
                                  # rename the germplasmName column to GID
                                  rename(GID=germplasmName)))

blups
```

Selection index:
```{r}
SIwts<-c(DM=15,
         #MCMDS=-10,
         logFYLD=20,
         logDYLD=20)
SIwts
```

Only difference: _do not_ subset the kinship matrix. Or more precisely, keep any genotypes meant to be either in the training set (phenotyped-and-genotyped) and those that are selection candidates (not-necessarily-genotyped).

In this example, simply leave all lines in the kinship matrix.

## Run genomic predictions
```{r}
gpreds<-runGenomicPredictions(modelType="A",
                              selInd=TRUE, SIwts=SIwts,
                              blups=blups,
                              grms=list(A=A),
                              ncores=3)
```

## Extract GEBV
Let's look at the output.
```{r}
gpreds
```
We have a single-row `tibble`. 

To access a simple table listing GEBV for each trait _and_ the selection index:
```{r}
gpreds$gblups[[1]]
```
At this point, you can use the **SELIND** predictions directly to rank and select parents. 

Example: sort by SELIND and pick the top 10...

```{r}
gpreds$gblups[[1]] %>% 
     arrange(desc(SELIND)) %>% 
     slice(1:10)
```
For more detailed output, including variance component estimates:
```{r}
gpreds$genomicPredOut[[1]]
```

```{r}
gpreds$genomicPredOut[[1]]$varcomps[[1]]
```

## Save the results
```{r}
saveRDS(gpreds,file = here::here("output","genomicPredictions.rds"))
```

